Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2018-06-15 articles/formalism/article.tex
updated abstract and section 4- polarized filaments

\abstract{
  \noindent 
  \revisit{We derive full-sky, real-space operators that convert between polarization Stokes $Q/U$ parameters to the coordinate-independent scalar $E$/$B$ modes that are widely used in Cosmic Microwave Background(CMB) and cosmic shear analysis.  The covolution kernels split naturally into angular and radial parts and we show explicitly how the radial extent of the convolution kernel depends on the targeted band-limit.  We show that an arbitrary radial dependence can produce $E/B$-like maps and that these are simply filtered versions of the standard $E/B$ maps.  This allows us to compute $E/B$ maps in real space with a compactly-supported kernel, an approach that can guarantee the avoidance of known foreground regions and can be employed in a massively parallel scheme at high-resolution.  We can recover the standard power spectrum of the polarized CMB sky by correcting the power spectra of these maps with a simple window function.  The spatial structure of the real space operator provides great intuition for the $E/B$ structure of polarized, filamentary galactic foregrounds and we predict non-zero B-mode signature that is expected from finite filaments in the sky.
    %
     We cast the standard CMB polarization analysis operators in a matrix-vector notation which facilitates the derivations and shows that the kernels relate directly to spin-0 $Y_{\ell 2}$ spherical harmonic functions.  This new notation also allows us to derive real space operators which decompose the measured Stokes parameters into their even and odd-parity parts, without ever evaluating the scalar $E$/$B$ fields themselves.  This paper is the first part in a series of papers that explore real-space computation of polarization modes and their applications.
  } 
  \\
  \\
%  We derive fully sky real space operators to translate Stokes Q \& U parameters to scalars $E$ \& $B$ and vice versa. We explicitly show that these local real space operator are fully characterized by the spin-0 $Y_{\ell 2}$ spherical harmonic functions. These real space operators make transparent the association of radial and tangential patterns of polarization with the E modes and that of  the clockwise and anti-clockwise pinwheel patterns of polarization with B-modes. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations. Using this new notation also allows us to derive real space operators which decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We use these analytical derivations us to quantify the non-local nature of the relation between the two representations of the polarization field. We present a prescription for generalizing these operators, which allow the non-local behavior of these operators to be a tunable parameter. We derive constraints on these generalized operators to ensure reliable recovery of the standard $E$ and $B$ fields and the corresponding power spectra. \revisit{This is part 1 of a series of papers where we have derived the real space operators. In the following papers we will explore some of their applications.} \comment{Needs better blending.}
}

>>>>

\abstract{
\noindent 
\revisit{We derive full-sky, real-space operators that convert between polarization Stokes $Q/U$ parameters to the coordinate-independent scalar $E$/$B$ modes that are widely used in Cosmic Microwave Background(CMB) and cosmic shear analysis. We also derive real space operators that decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations.  For all these real space operators we show that the kernels split naturally into angular and radial parts and we show explicitly how the radial extent of these kernels depends on the targeted band-limit. We show that the kernels can be interpreted either as a complex convolving beam or the Green's function when they are expressed as functions of the forward or inverse rotation Euler angles respectively. We show that an arbitrary radial dependence, satisfying the constraint that it vanishes at angular separations of $0$ and $\pi$, can produce $E/B$-like maps and that these are simply filtered versions of the standard $E/B$ maps.  This allows us to compute $E/B$ maps in real space with a compactly-supported kernel, an approach that can guarantee the avoidance of known foreground regions and can be employed in a massively parallel scheme at high-resolution. We show that the spin raising $\eth^2$ and lowering $\bar{\eth}^2$ operators are a special case of these generalized operators and present their band limited versions. We can recover the standard power spectrum of the polarized CMB sky by correcting the power spectra of these maps with a simple window function. We present a scheme to derive this window function for any radial dependence. The spatial structure of the real space operator provides great intuition for the $E/B$ structure of polarized, filamentary galactic foregrounds and we predict non-zero B-mode signature that is expected from finite filaments in the sky. This paper is the first part in a series of papers that explore real-space computation of polarization modes and their applications.
  } 
  \\
  \\
%  We derive fully sky real space operators to translate Stokes Q \& U parameters to scalars $E$ \& $B$ and vice versa. We explicitly show that these local real space operator are fully characterized by the spin-0 $Y_{\ell 2}$ spherical harmonic functions. These real space operators make transparent the association of radial and tangential patterns of polarization with the E modes and that of  the clockwise and anti-clockwise pinwheel patterns of polarization with B-modes. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations. Using this new notation also allows us to derive real space operators which decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We use these analytical derivations us to quantify the non-local nature of the relation between the two representations of the polarization field. We present a prescription for generalizing these operators, which allow the non-local behavior of these operators to be a tunable parameter. We derive constraints on these generalized operators to ensure reliable recovery of the standard $E$ and $B$ fields and the corresponding power spectra. \revisit{This is part 1 of a series of papers where we have derived the real space operators. In the following papers we will explore some of their applications.} \comment{Needs better blending.}
}

### 2018-06-15 articles/formalism/sections/discussion.tex
updated abstract and section 4- polarized filaments

These real space operators provide a spatially intuitive way of understanding the different decompositions of the Stokes vector. We explicitly  demonstrated that all the real space operators have the common characteristic  of being separable into a azimuthal part that is band limit independent and radial part that incorporates all the band limit dependence. The azimuthal part of the operator is primarily responsible for the requisite spinorial decomposition, while the radial weights determine the non-local dependence of the construction of the resultant fields on the original fields. Using the self similarity property of the radial functions we define a non-locality parameter and argued that it is proportional to $\ell_{\rm max}^{-1}$. The scalar decomposition of the Stokes parameters being of primary interest,  we empirically derived the non-locality parameter $\beta_0$ for this operation and showed that the proportionality constant $\ell_{0}=22$ provided a good prediction for the angular distance at which the ${}_{\mm}f$ falls below 1\% of its maxima as seen in \fig{fig:rad_ker_decay}.

Our careful study of the real space operators reveals the dual interpretation of the operators as either the Green's function or a convolving beam depending on whether the kernels are expressed in terms of the forward rotation Euler angles or the inverse rotation Euler angles. While the convolution interpretation is a familiar one, the Green's function interpretation is a new way of looking at this operation. This new interpretation allows us to think of ${}_{+2}X$ as some spin-2 charge which radiates out a complex spin-0 scalar field $E+iB$. The resultant complex scalar maps can be then understood as arising from superposition of the radiating field emanating from all the spin charges on the sphere. The $E/B$ mode maps are merely the real and imaginary parts of this resultant complex scalar field.

>>>>

These real space operators provide a spatially intuitive way of understanding the different decompositions of the Stokes vector on the sphere. We explicitly  demonstrated that all the real space operators have the common characteristic  of being separable into a azimuthal part that is band limit independent and radial part that incorporates all the band limit dependence. The azimuthal part of the operator is primarily responsible for the requisite spinorial decomposition, while the radial weights determine the non-local dependence of the construction of the resultant fields on the original fields. Using the self similarity property of the radial functions we define a non-locality parameter and argued that it is proportional to $\ell_{\rm max}^{-1}$. The scalar decomposition of the Stokes parameters being of primary interest,  we empirically derived the non-locality parameter $\beta_0$ for this operation and showed that the proportionality constant $\ell_{0}=22$ provided a good prediction for the angular distance at which the ${}_{\mm}f$ falls below 1\% of its maxima as seen in \fig{fig:rad_ker_decay}.

Our careful study of the real space operators reveals the dual interpretation of the operators as either the Green's function or a complex convolving beam depending on whether the kernels are expressed in terms of the forward rotation Euler angles or the inverse rotation Euler angles. While the convolution interpretation is a familiar one, the Green's function interpretation is a new way of looking at this operation. This new interpretation allows us to think of ${}_{+2}X$ as some spin-2 charge which radiates out a complex spin-0 scalar field $E+iB$. The resultant complex scalar maps can be then understood as arising from superposition of the radiating field emanating from all the spin charges on the sphere. The $E/B$ mode maps are merely the real and imaginary parts of this resultant complex scalar field.

### 2018-06-15 articles/formalism/sections/kernel_paintbrush.tex
updated abstract and section 4- polarized filaments

%\revisit{A stacking analysis of Planck data \citep{2016A&A...586A.141P} sees $E>0$ along filaments (selected from intensity data), but no $B$-mode signal is above the noise.  We predict that it should be there in higher fidelity data.}
\revisit{A stacking analysis of Planck data \citep{2016A&A...586A.141P} sees $E>0$ along filaments (selected from intensity data), but claim no $B$-mode signal. We predict that a B-mode signal from filaments should be present. A naive filament stacking analysis is only suited for making E-mode detections. Detecting B-modes from filaments requires a more careful edge stacked filaments analysis. While the detectability of B-mode signature from filaments from Planck data calls for a more careful assessment, it should definitely be detectable in higher fidelity data.}

The intuition from the real-space kernels holds also when we distort the shape of the filament.  If the filament were bent around into a circle, the positive and negative parts of the $B$ pattern will cancel, and we are left with a hoop of pure $E$ pattern (note that this cancellation would not happen for an ellipse or any loop of non-contant radius of curvature).  The same general description holds for a spiral-shaped filament, which can can be viewed as distortion of the straight filament.  The filament is highlighted by positive $E>0$.  The $E$-pattern is more negative on the interior of a curve than on the exterior, and the concentric rings of filamentary structure make an increasingly negative $E$ value inside.  The $B$-pattern is again concentrated at the ends of the filament in an oriented pair of positive/negative fluctuations. \comment{Add that the spiral has a constant radius of curvature and thats the reason you dont see B-mode signal in intermediate regions of the spiral!}

\revisit{Note that we have made the simplifying assumption that the polarization headless vectors are perpendicular to the tangent to the filament at any given point and this results in dominantly generating an E-mode signal while the B-mode signal is present only near the edges. By similar arguments it can be seen that that polarization pattern will be dominated by B-modes if the polarization vector is always at $+45$ (or $-45$) degrees to the tangent to the filament while the E-mode signal would only be present at the edges. Therefore a careful study of the $E/B$ mode power in a filament can provide insights into the orientation of the magnetic field with respect to the length of the filament, assuming the polarization vectors (headless) are always perpendicular to the magnetic field vector.}

>>>>

%\revisit{A stacking analysis of Planck data \citep{2016A&A...586A.141P} sees $E>0$ along filaments (selected from intensity data), but no $B$-mode signal is above the noise.  We predict that it should be there in higher fidelity data.}
\revisit{A stacking analysis of Planck data \citep{2016A&A...586A.141P} sees $E>0$ along filaments (selected from intensity data), but claim no $B$-mode signal. We predict that a B-mode signal from filaments should be present. A naive filament stacking analysis is only suited for making E-mode detections. Detecting B-modes from filaments requires a more careful edge stacked filaments analysis. While the detectability of B-mode signature from filaments from Planck data calls for a more careful assessment, it should be detectable in higher fidelity data.}

The intuition from the real-space kernels holds also when we distort the shape of the filament.  If the filament were bent around into a circle, the positive and negative parts of the $B$ pattern will cancel, and we are left with a hoop of pure $E$ pattern (note that this cancellation would not happen for an ellipse or any loop of non-contant radius of curvature).  The same general description holds for a spiral-shaped filament, which can can be viewed as distortion of the straight filament.  The filament is highlighted by positive $E>0$.  The $E$-pattern is more negative on the interior of a curve than on the exterior, and the concentric rings of filamentary structure make an increasingly negative $E$ value inside.  The $B$-pattern is again concentrated at the ends of the filament in an oriented pair of positive/negative fluctuations. \revisit{The spiral filament showcased in \fig{fig:polfilaments} has a constant radius of curvature and thats the reason for not seeing a B-mode signal in intermediate regions of the spiral. For filaments that don't have a constant radius of curvature one expects to generate more B-mode power.} \comment{This is an important point right? So even if the magnetic field is perfectly aligned with the filament, one can generate B-mode signal in intermediate regions of the filament just by having the filament with changing radius of curvature. There is no reason for filaments to have a constant radius of curvature.}

\revisit{Note that we have made the simplifying assumption that the polarization headless vectors are perpendicular to the magnetic field which is always assumed parallel to the tangent to the filament at any given point and this results in generating dominantly an E-mode signal while the B-mode signal is present only near the edges. By similar arguments it can be shown that that polarization pattern will be dominated by B-modes if the polarization vector is aligned at $+45$ (or $-45$) degrees to the tangent to the filament while the E-mode signal would only be present at the edges. Therefore a careful study of the $E/B$ mode power in filaments can provide insights into the orientation of the magnetic field with respect to the length of the filament.}

### 2018-06-15 articles/formalism/sections/real_space_formalism.tex
updated abstract and section 4- polarized filaments

\subsection{Matrix notation} \label{sec:mat_pol_intro}
Our derivation of real space operators is more transparent in a matrix-vector notation.\footnote{While we work with the matrix and vector sizes given in terms of some pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}
We introduce a matrix that encodes spin spherical harmonic basis vectors:
%
\beq
{}_{|s|}\mathcal{Y}= \bmat _{+s}Y & 0 \\ 0 & _{-s}Y \emat _{2 \rm N_{\rm pix} \times 2 \rm N_{\rm alms}} ;\qquad  {}_{|s|}\mathcal{Y}^{\ddagger}= \Delta \Omega \bmat _{+s}Y^{\dagger} & 0 \\ 0 & _{-s}Y^{\dagger} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm pix}} \,,
\eeq
%
where $s$ denotes the spin of the basis functions and our \revisit{definition of ${}_{|s|}\mathcal{Y}^{\ddagger}$ differs from the conventional conjugate transpose operation by the factor $\Delta \Omega$.  We introduce this to ensure the ortho-\textit{normality} of these operations, so that ${}_{|s|}\mathcal{Y}^{\ddagger} {}_{|s|}\mathcal{Y} = I_{2 \rm N_{\rm alms} \times2 \rm N_{\rm alms}}$ on the discretized sphere, and maintain the standard definition of spherical harmonics. [Orthonormal only in infimetesimal pixel limit?]}

@@

The operator $\mathcal{I}$ is Hermitian and is a band limited version of the delta function owing to the identity: $\lim_{\ell_{\rm max} \rightarrow \infty} \mathcal{I} = \delta(\hat{n}_i - \hat{n}_j)$. For all practical purposes $\mathcal{I}$ acts like an identity operator as ascertained by the following set of identities: (i) $\mathcal{I} \mathcal{I}=\mathcal{I}$ ; (ii) $\mathcal{D} \mathcal{I}=\mathcal{D}$. $\mathcal{D}$ is a complex but symmetric matrix and $\mathcal{D}^*$ is its inverse in this band limited sense: $\mathcal{D}^* \mathcal{D}=\mathcal{I}$. Using these properties\footnote{While testing the real space operator identities one encounters terms like $\mathcal{D} \mathcal{I}^*,\mathcal{I}^*\mathcal{I}$ and $\mathcal{I}\mathcal{I}^*$ which cannot be simply interpreted but they always occur in pairs with opposite signs that exactly cancel each other.} of the operators $\mathcal{I}$ and $\mathcal{D}$ , one can verify that the real space operators satisfy the following identities:
%
\begin{subequations}
\beqry
\bar O_E \, \bar O_E &=& \bar O_E;\qquad \bar O_B \, \bar O_B = \bar O_B \,, \\
\bar O_E \, \bar O_B &=& 0 \,,\\
\bar O_E + \bar O_B &=& \mathcal{I} \,,
\eeqry
\end{subequations}
%
which are the real space analogues of their harmonic space counterparts discussed in \sec{sec:mat_pol_intro}. Thus they are exactly orthogonal and idempotent. Note that unlike in the harmonic case, the sum of the operators is the band limited identity operator $\mathcal{I}$. This non-exactness is representative of the loss of information resulting from making this transformation on measured data with some imposed band limit. Forcing the sum of the operators to be exactly an identity matrix compromises the orthogonality property of the $\bar{O}_E$ \& $\bar{O}_B$ operators which \revisit{is exact and a more crucial property of the operators.[K: what does this mean?]}

>>>>

\subsection{Matrix notation} \label{sec:mat_pol_intro}
Our derivation of real space operators is more transparent in a matrix-vector notation.\footnote{While we work with the matrix and vector sizes given in terms of some pixelization parameter $\rm N_{\rm pix}$, all the relations are equally valid in the continuum limit attained by allowing $\rm N_{\rm pix}\rightarrow \infty$}
We introduce a matrix that encodes spin spherical harmonic basis vectors:
%
\beq
{}_{|s|}\mathcal{Y}= \bmat _{+s}Y & 0 \\ 0 & _{-s}Y \emat _{2 \rm N_{\rm pix} \times 2 \rm N_{\rm alms}} ;\qquad  {}_{|s|}\mathcal{Y}^{\ddagger}= \Delta \Omega \bmat _{+s}Y^{\dagger} & 0 \\ 0 & _{-s}Y^{\dagger} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm pix}} \,,
\eeq
%
where $s$ denotes the spin of the basis functions and our \revisit{definition of ${}_{|s|}\mathcal{Y}^{\ddagger}$ differs from the conventional conjugate transpose operation by the factor $\Delta \Omega$.  We introduce this to ensure the ortho-\textit{normality} of these operations, so that ${}_{|s|}\mathcal{Y}^{\ddagger} {}_{|s|}\mathcal{Y} = I_{2 \rm N_{\rm alms} \times2 \rm N_{\rm alms}}$ on the discretized sphere, and maintain the standard definition of spherical harmonics.} \comment{[Orthonormal only in infimetesimal pixel limit?, I think it doesnt have to be infitesimal but sufficiently small.]}

@@

The operator $\mathcal{I}$ is Hermitian and is a band limited version of the delta function owing to the identity: $\lim_{\ell_{\rm max} \rightarrow \infty} \mathcal{I} = \delta(\hat{n}_i - \hat{n}_j)$. For all practical purposes $\mathcal{I}$ acts like an identity operator as ascertained by the following set of identities: (i) $\mathcal{I} \mathcal{I}=\mathcal{I}$ ; (ii) $\mathcal{D} \mathcal{I}=\mathcal{D}$. $\mathcal{D}$ is a complex but symmetric matrix and $\mathcal{D}^*$ is its inverse in this band limited sense: $\mathcal{D}^* \mathcal{D}=\mathcal{I}$. Using these properties\footnote{While testing the real space operator identities one encounters terms like $\mathcal{D} \mathcal{I}^*,\mathcal{I}^*\mathcal{I}$ and $\mathcal{I}\mathcal{I}^*$ which cannot be simply interpreted but they always occur in pairs with opposite signs that exactly cancel each other.} of the operators $\mathcal{I}$ and $\mathcal{D}$ , one can verify that the real space operators satisfy the following identities:
%
\begin{subequations}
\beqry
\bar O_E \, \bar O_E &=& \bar O_E;\qquad \bar O_B \, \bar O_B = \bar O_B \,, \\
\bar O_E \, \bar O_B &=& 0 \,,\\
\bar O_E + \bar O_B &=& \mathcal{I} \,,
\eeqry
\end{subequations}
%
which are the real space analogues of their harmonic space counterparts discussed in \sec{sec:mat_pol_intro}. Thus they are exactly orthogonal and idempotent. Note that unlike in the harmonic case, the sum of the operators is the band limited identity operator $\mathcal{I}$. This non-exactness is representative of the loss of information resulting from making this transformation on measured data with some imposed band limit. Forcing the sum of the operators to be exactly an identity matrix compromises the orthogonality property of the $\bar{O}_E$ \& $\bar{O}_B$ operators which \revisit{is exact and a more crucial property of the operators.} \comment{[K: what does this mean?, A: Just mean that power conservation is not as important as the orthogonality of the E/B mode functions.]}
